Dynamics in Algorithmic Recourse

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.qmd
more work on abstract

This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually impemented by a proportion of individuals and the classifier is updated in response. @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift. As this game of implementing AR and updating the classifier is repeated, individuals who receive and implement algorithmic recourse end up forming a distinct subgroup inside the target class, which may leave them vulnerable to discrimination. Through future experiments we want to investigate if this phenomenon is robust across different benchmark datasets and counterfactual generators.

@@

Alongside my research I have developed open-source implementations related to explainable AI. [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/) is a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages like Python and R. I have recently submitted the package along with a companion paper as a proposal for a main talk at [JuliaCon](https://juliacon.org/2022/). [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/) is a small Julia package that can be used to recover Bayesian representations of deep neural networks through Laplace approximation in a post-hoc manner. It is inspired by a recent paper [@daxberger2021laplace] and has also been submitted to JuliaCon. Finally, [deepvars](https://github.com/pat-alt/deepvars) is an R package that implements an approach towards vector autoregression that leverages deep learning. This was originally my master's thesis and later presented at the NeurIPS 2021 MLECON workshop. I have also published several blog posts on explainable AI and probabilisitic ML in an effort to make my research accessible to a broad audience.

### Probabilistic methods for realistic counterfactual explanations

Probabilistic machine learning can be leveraged in this context and more generally facilitates inference and interpretability. It is also closely related to Bayesian statistics, which has played an important role in both finance and economics for many years.

To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not. To quantify their joint likelihood, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions. While the former approach is more versatile since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient. In my work I want to explore how recent advances in post-hoc uncertainty quantification can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.

### CE for time series

Data sets in finance and economics typically involve time series data. Therefore, I am naturally interested in the application of explainable AI to sequential data, an area which has so far not been explored extensively. In the future, I want to work on counterfactual explanations for time series models. I am also interested in seeing if and how Laplace approximation can be used for Bayesian deep learning with time series data. I hope that the findings from both of these projects can ultimately be used to build complex but interpretable time series models for classification and forecasting in finance and economics.

### Explainable black-box models for time series

For example, I would like to leverage effortless Bayesian deep learning to make our proposed Deep Vector Autoregression model explainable.

>>>>

This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually impemented by a proportion of individuals and the classifier is updated in response. @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift. As this game of implementing AR and updating the classifier is repeated, the decision boundary moves in the opposite direction of the original training samples in the target class. We consider several aspects of these dynamics as problematic. Firstly, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as credit worthy. Average default risk across all borrowers can therefore be expected to increase. Conversely, lenders that anticipate or detect these sorts of dynamics may choose to still deny credit to individuals that have implemented AR, thereby comprimising the validity of AR.

To the best of my knowledge this is the first work investigating endogenous dynamic in algorithmic recourse. In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets (including German credit [@germancredit1994], Boston Housing [@boston1993] and COMPAS [@compas2016]). Furthermore, I want to asses to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator. To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.

@@

Until recently there existed only one open-source library that provides a unifying approach to generate and benchmark counterfactual explanations for models built and trained in Python [@pawelczyk2021carla]. To address this limitation I have developed [CounterfactualExplanations.jl](https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/): a Julia package that can be used to generate counterfactual explanations for models developed and trained not only in Julia, but also in other popular programming languages like Python and R. The package and companion paper are currently pending acceptance for a main talk at [JuliaCon](https://juliacon.org/2022/).

### Probabilistic Methods for Realistic CE

To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not. To quantify their joint likelihood, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions. While the former approach is more versatile since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient. The approach proposed by @schut2021generating and used to generate the examples in @fig-dynamics falls into the latter category. In future work I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]

### XAI for Time Series Models

Datasets in Finance and Economics often involve autoregressive time series, which is modelled differently than static, cross-sectional data. One popular approach to macroeconomic forecasting, for example, is Vector Autoregression (VAR), which involves an interpretable system of linear equations. Recent work has shown that machine learning can be leveraged in this context to improve forecasting performance, albeit at the sacrifice of explainability [@altmeyer2021deep].^[For an open-source implementation of our approach see [deepvars](https://github.com/pat-alt/deepvars): an R package Deep Vector Autoregression]. Through my PhD research I hope to ultimately adress that shortfall through explainable AI and probabilistic methods.

#### Counterfactual Explanations

I would like to investigate if and how CE can be used to leverage machine learning for time series forecasting without sacrificing explainability. In the context of conventional VAR counterfactual outcomes are actually well-studied [@kilian2017structural], which makes macroeconomic forecasting and interesting case study. More generally though, work on applying CE to black-box models for time series data has so far been limited and therefore warrants attention.

#### Probabilistic Methods

I am also interested in seeing if and how Laplace Redux can be applied to reccurent neural networks.

### 2022-04-25 dev/submissions/aies/extended_abstract/extended_abstract.tex
more work on abstract

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

This project investigates \textbf{endogenous} domain and model shifts,
that is shifts that occur when AR is actually impemented by a proportion
of individuals and the classifier is updated in response.
Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
involving a probabilistic classifier and a greedy counterfactual
generator proposed by Schut et al. (2021): AR leads to a domain shift,
which in turn causes a drastic model shift. As this game of implementing
AR and updating the classifier is repeated, individuals who receive and
implement algorithmic recourse end up forming a distinct subgroup inside
the target class, which may leave them vulnerable to discrimination.
Through future experiments we want to investigate if this phenomenon is
robust across different benchmark datasets and counterfactual
generators.

@@

Alongside my research I have developed open-source implementations
related to explainable AI.
\href{https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/}{CounterfactualExplanations.jl}
is a Julia package that can be used to generate counterfactual
explanations for models developed and trained not only in Julia, but
also in other popular programming languages like Python and R. I have
recently submitted the package along with a companion paper as a
proposal for a main talk at \href{https://juliacon.org/2022/}{JuliaCon}.
\href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}
is a small Julia package that can be used to recover Bayesian
representations of deep neural networks through Laplace approximation in
a post-hoc manner. It is inspired by a recent paper (Daxberger et al.
2021) and has also been submitted to JuliaCon. Finally,
\href{https://github.com/pat-alt/deepvars}{deepvars} is an R package
that implements an approach towards vector autoregression that leverages
deep learning. This was originally my master's thesis and later
presented at the NeurIPS 2021 MLECON workshop. I have also published
several blog posts on explainable AI and probabilisitic ML in an effort
to make my research accessible to a broad audience.

\hypertarget{probabilistic-methods-for-realistic-counterfactual-explanations}{%
\subsection{Probabilistic methods for realistic counterfactual
explanations}\label{probabilistic-methods-for-realistic-counterfactual-explanations}}

Probabilistic machine learning can be leveraged in this context and more
generally facilitates inference and interpretability. It is also closely
related to Bayesian statistics, which has played an important role in
both finance and economics for many years.

To ensure that the generated explanations are realistic it is important
to understand which input-output pairs are likely and which are not. To
quantify their joint likelihood, previous work has either relied on
generative models or restricted the analysis to probabilistic models
that incorporate uncertainty in their predictions. While the former
approach is more versatile since it is applicable to both deterministic
and probabilistic models, the latter is computationally much more
efficient. In my work I want to explore how recent advances in post-hoc
uncertainty quantification can be leveraged to generate realistic and
unambiguous counterfactual explanations for any model.

\hypertarget{ce-for-time-series}{%
\subsection{CE for time series}\label{ce-for-time-series}}

Data sets in finance and economics typically involve time series data.
Therefore, I am naturally interested in the application of explainable
AI to sequential data, an area which has so far not been explored
extensively. In the future, I want to work on counterfactual
explanations for time series models. I am also interested in seeing if
and how Laplace approximation can be used for Bayesian deep learning
with time series data. I hope that the findings from both of these
projects can ultimately be used to build complex but interpretable time
series models for classification and forecasting in finance and
economics.

\hypertarget{explainable-black-box-models-for-time-series}{%
\subsection{Explainable black-box models for time
series}\label{explainable-black-box-models-for-time-series}}

For example, I would like to leverage effortless Bayesian deep learning
to make our proposed Deep Vector Autoregression model explainable.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
Miguel Hernández-Lobato. 2020. {``Getting a Clue: A Method for
Explaining Uncertainty Estimates.''} \emph{arXiv Preprint
arXiv:2006.06848}.

@@

@@

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

This project investigates \textbf{endogenous} domain and model shifts,
that is shifts that occur when AR is actually impemented by a proportion
of individuals and the classifier is updated in response.
Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
involving a probabilistic classifier and a greedy counterfactual
generator proposed by Schut et al. (2021): AR leads to a domain shift,
which in turn causes a drastic model shift. As this game of implementing
AR and updating the classifier is repeated, the decision boundary moves
in the opposite direction of the original training samples in the target
class. We consider several aspects of these dynamics as problematic.
Firstly, as the decision boundary moves in the direction of the
non-target class, counterfactual paths become shorter: in the loan
example, individuals that previously would have been denied credit based
on their input features are suddenly considered as credit worthy.
Average default risk across all borrowers can therefore be expected to
increase. Conversely, lenders that anticipate or detect these sorts of
dynamics may choose to still deny credit to individuals that have
implemented AR, thereby comprimising the validity of AR.

To the best of my knowledge this is the first work investigating
endogenous dynamic in algorithmic recourse. In future experiments I want
to investigate how this phenomenon plays out across different benchmark
datasets (including German credit ({``German Credit Data''} 1994),
Boston Housing ({``Boston Housing Data''} 1993) and COMPAS (Larson et
al. 2016)). Furthermore, I want to asses to what extent the magnitude
and direction of domain and model shifts depends on the choice of the
counterfactual generator. To this end, I am currently supervising a
group of undergraduate students, who are tackling some of these tasks in
their final-year research project.

@@

Until recently there existed only one open-source library that provides
a unifying approach to generate and benchmark counterfactual
explanations for models built and trained in Python (Pawelczyk et al.
2021). To address this limitation I have developed
\href{https://www.paltmeyer.com/CounterfactualExplanations.jl/stable/}{CounterfactualExplanations.jl}:
a Julia package that can be used to generate counterfactual explanations
for models developed and trained not only in Julia, but also in other
popular programming languages like Python and R. The package and
companion paper are currently pending acceptance for a main talk at
\href{https://juliacon.org/2022/}{JuliaCon}.

\hypertarget{probabilistic-methods-for-realistic-ce}{%
\subsection{Probabilistic Methods for Realistic
CE}\label{probabilistic-methods-for-realistic-ce}}

To ensure that the generated explanations are realistic it is important
to understand which input-output pairs are likely and which are not. To
quantify their joint likelihood, previous work has either relied on
generative models or restricted the analysis to probabilistic models
that incorporate uncertainty in their predictions. While the former
approach is more versatile since it is applicable to both deterministic
and probabilistic models, the latter is computationally much more
efficient. The approach proposed by Schut et al. (2021) and used to
generate the examples in Figure~\ref{fig-dynamics} falls into the latter
category. In future work I want to explore how recent advances in
post-hoc uncertainty quantification, most notably Laplace Redux
(Daxberger et al. 2021), can be leveraged to generate realistic and
unambiguous counterfactual explanations for any model.\footnote{For some
  initial work on this see my Julia implementation of Laplace Redux:
  \href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}.}

\hypertarget{xai-for-time-series-models}{%
\subsection{XAI for time series
models}\label{xai-for-time-series-models}}

Datasets in Finance and Economics often involve autoregressive time
series, which is modelled differently than static, cross-sectional data.
One popular approach to macroeconomic forecasting, for example, is
Vector Autoregression (VAR), which involves an interpretable system of
linear equations. Recent work has shown that machine learning can be
leveraged in this context to improve forecasting performance, albeit at
the sacrifice of explainability (Altmeyer, Agusti, and Vidal-Quadras
Costa 2021).\footnote{For an open-source implementation of our approach
  see \href{https://github.com/pat-alt/deepvars}{deepvars}: an R package
  Deep Vector Autoregression}. Through my PhD research I hope to
ultimately adress that shortfall through explainable AI and
probabilistic methods.

\hypertarget{counterfactual-explanations}{%
\subsubsection{Counterfactual
Explanations}\label{counterfactual-explanations}}

I would like to investigate if and how CE can be used to leverage
machine learning for time series forecasting without sacrificing
explainability. In the context of conventional VAR counterfactual
outcomes are actually well-studied (Kilian and Lütkepohl 2017), which
makes macroeconomic forecasting and interesting case study. More
generally though, work on applying CE to black-box models for time
series data has so far been limited and therefore warrants attention.

\hypertarget{probabilistic-methods}{%
\subsubsection{Probabilistic Methods}\label{probabilistic-methods}}

I am also interested in seeing if and how Laplace approximation can be
used for Bayesian deep learning with time series data. I hope that the
findings from both of these projects can ultimately be used to build
complex but interpretable time series models for classification and
forecasting in finance and economics.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2021deep}{}}%
Altmeyer, Patrick, Marc Agusti, and Ignacio Vidal-Quadras Costa. 2021.
{``Deep Vector Autoregression for Macroeconomic Data.''}
\url{https://thevoice.bse.eu/wp-content/uploads/2021/07/ds21-project-agusti-et-al.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
Miguel Hernández-Lobato. 2020. {``Getting a Clue: A Method for
Explaining Uncertainty Estimates.''} \emph{arXiv Preprint
arXiv:2006.06848}.

\leavevmode\vadjust pre{\hypertarget{ref-boston1993}{}}%
{``Boston Housing Data.''} 1993.
\url{https://archive.ics.uci.edu/ml/machine-learning-databases/housing/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
{``German Credit Data.''} 1994.
\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kilian2017structural}{}}%
Kilian, Lutz, and Helmut Lütkepohl. 2017. \emph{Structural Vector
Autoregressive Analysis}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-compas2016}{}}%
Larson, Jeff, Surya Mattu, Lauren Kirchner, and Julia Angwin. 2016.
{``COMPAS.''} \url{https://github.com/propublica/compas-analysis}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias
Richter, and Gjergji Kasneci. 2021. {``Carla: A Python Library to
Benchmark Algorithmic Recourse and Counterfactual Explanation
Algorithms.''} \emph{arXiv Preprint arXiv:2108.00783}.
